---
output: html_document
---


The following examples illustrate the preprocessing of spatial data sets using
the sp and raster package.

As always, we start with importing the packages and setting the working
directory.
```{r, echo=FALSE, warning=FALSE, message=FALSE, results="hide"}
library(raster)
library(sp)
library(rgdal)  # Necessary to read vector files when reading with sp

setwd("D:/active/moc/dm/examples/data_procd/spatial_data/")
```

#####  Vector data sets I/O
A standard exchange format for vector data sets is the ESRI shape format. The
shape "file" is actually not a single file but a group of files with the same
name but different extensions. To load a vector data set into R, the name of
the actual shape file (i.e. extension "shp") has to be supplied. In addition,
the layer name must be given to the function. In 99.99% of all cases, this 
layer name is just the file name.

For reading a vector data set, the `readOGR` function of the sp package is used:
```{r}
print(getwd())
vector <- readOGR("data_2014_subset1.shp", layer = "data_2014_subset1")
```
As one can see from the short information return of the function, the example file
was of type polygon with 161 points (i.e. features) and 6 attribute values per
point.

What you get from readOGR is some kind of spatial data frame. Since this is a
point vector, it will be a spatial point data frame:
```{r}
class(vector)
```

If projection information has been supplied to the shape file (i.e. if a consistent
prj file is part of the data set), the projection information has already been
stored in the spatial data frame. This can be checked (or accessed) by using the
`projection` function:
```{r}
projection(vector)
```

If you want to write a spatial vector data frame to your disk, just use the
`writeOGR` function:
```{r}
writeOGR(vector, "new_data_set.shp", layer = "new_data_set", 
         driver = "ESRI Shapefile", overwrite_layer=FALSE)
```
As driver, any of the GDAL drivers can be used. Personally, I vote for the default
of attribute overwrite_layer which is FALSE (and hence you can neglect it) in
order to ensure that existing data is not overwritten accidentially.


#####  Raster data set I/O
A standard exchange format for raster data sets is the GeoTiff format. As for 
other GIS data formats, it supplies meta information on its corner coordinates and
also its reference system. 
```{r}
raster <- raster("LC82100502014328LGN00_B5.tif")
projection(raster)
```
To write a raster, use the `writeRaster` function:
```{r}
writeRaster(raster, "new_raster.tif", "GTiff", overwrite = FALSE)
```


#####  Multiple raster data sets I/O
Many task require the reading of multiple raster data sets. Assuming that they
have all the same projection, extend and resolution, individual rasters can 
be combined to stacks using the `stack` function. The combination can be based
on individual rasters readily available in memory or raster data sets which
are not loaded (but will be as part of the function). This function can also handle
the i/o of multi-layer raster files which is common for RGB composits.
```{r}
landsat <- stack(raster, "LC82100502014328LGN00_B1.tif", 
                 "LC82100502014328LGN00_B2.tif", "LC82100502014328LGN00_B3.tif")
projection(landsat)
```
Tip: the `resample` or `projectRaster` function can help you in achiving that
all raster layers have the same projection and internal geometry.


##### XYZ data I/O
If geo-datasets do not come as vectors or rasters but as csv files, it is no
problem to convert the resulting data frame from the `read.table` function to
a spatial data frame as long as you can supply an x (i.e. easing) and y (i.e.
northing) coordinate for each row of the data frame.

In the following example, the x and y coordinates are already part of the csv
file.
```{r}
xyz <- read.table("plots_veg_anm_geo_2014.csv", header=TRUE, sep=",")
```

Since this is a plain csv file, there is no projection information available.
```{r}
projection(xyz)
```
In order to convert the data frame to a spatial data frame, we first have to
supply easting and northing coordinates for each row. In the example at hand,
the coordinates are given in columns "Lon" and "Lat".
```{r}
coordinates(xyz) <- ~Lon+Lat
```
Finally, the projection has to be defined. In this case, the data comes from a GPS
device with standard latitude and longitude (i.e. WGS84) settings. The EPSG code
for that is 4326 and the proj4 string is as follows:
```{r}
projection(xyz) <- "+proj=longlat +ellps=WGS84 +datum=WGS84 +no_defs"
class(xyz)
```
Now, the spatial points data frame can be handled as any other.

Of course, the other way round is also possible for any kind of spatial vector
data frame:
```{r}
xyz <- as.data.frame(xyz)
class(xyz)
```
The resulting data frame has two additional columns x and y which contain the
easting and northing positions. These will always be added to the data frame.