# Data Handling {#data-handling}

One thing that most people fail to acknowledge is that visualizing data in R (or any other programming language for that matter) usually involves a little more effort than simply calling some plot function to create a meaningful graph. Data visualization is in essence an abstract representation of raw data. Most plotting routines available in R, however, are not designed to provide any useful data abstraction. This means that it is up to us to prepare our data to a level of abstraction that is feasible for what we want to show with our visualization. 

Therefore, before we start to produce plots, we will need to spend some time and effort to get familiar with some tools to manipulate our raw data sets. In particular, we will learn how to `subset()`, `aggregate()`, `sort()`, and `merge()` our data sets. For the sake of reproducibility, this workshop will make use of the `diamonds` data set (which comes with **ggplot2**) in all the provided examples.

```{r global settings, tidy = FALSE, message = FALSE, results = 'hide'}
### here's a rather handy way of loading all packages that you need
### for your code to work in one go
pkg <- c('ggplot2', 'latticeExtra', 'gridExtra', 'MASS', 
         'colorspace', 'plyr', 'Hmisc', 'scales')
jnk <- sapply(pkg, library, character.only = TRUE)

### load the diamonds data set (comes with ggplot2)
data(diamonds)
```

Right, enough of that introductory talk, let's start getting our hands dirty!


## Subsetting Data {#subset}

The `diamonds` data set from **ggplot2** is structured as follows:

```{r str, tidy = FALSE}
str(diamonds)
```

The `str()` command is probably the most useful command in all of R. It shows the complete structure of our data set and provides a 'road map' of how to access certain parts of the data.   

For example, the console output tells us that `diamonds$carat` is a numerical vector of length `r nrow(diamonds)`, whereas `diamonds$cut` is a factor with the ordered levels 

```{r levels}
levels(diamonds$cut)
```

Suppose we're a stingy person and don't want to spend too much money on the wedding ring for our loved one, we could create a data set only including diamonds that cost less than 1000 USD (though 1000 USD does still seem very generous to me).

```{r subset cheap, tidy = FALSE}
diamonds_cheap <- subset(diamonds, price < 1000)
```

Then our new data set would look like this

```{r str subset cheap, tidy = FALSE}
str(diamonds_cheap)
```

Now the new ```diamonds_cheap``` subset is a reduced version of the original ```diamonds``` data set only having `r nrow(diamonds_cheap)` entries instead of the original `r nrow(diamonds)` entries.

In case we were interested in a subset only including all diamonds of 'Premium' quality (column `cut`), the command would be

```{r subset premium, tidy = FALSE}
diamonds_premium <- subset(diamonds, cut == "Premium")
str(diamonds_premium)
```

Note the _two_ equal signs in order to specify our selection. This stems from an effort to be consistent with selection criteria such as "smaller than" (`<=`) or "not equal" (`!=`) and basically translates to "is equal"!

Any combinations of these conditional statements are valid within one and the same `subset` call, e.g.

```{r subset premium and cheap, tidy = FALSE}
diamonds_premium_and_cheap <- subset(diamonds, cut == "Premium" & 
                                       price <= 1000)
```

produces a rather strict subset only allowing diamonds of premium quality that cost less than 1000 USD.

In case we want _ANY_ of these, meaning all diamonds of premium quality _OR_ cheaper than 1000 USD, we would use the `|` operator to combine the two specifications:

```{r subset premium or cheap, tidy = FALSE}
diamonds_premium_or_cheap <- subset(diamonds, cut == "Premium" | 
                                      price <= 1000)
```

The _OR_ specification is much less rigid than the _AND_ specification which will result in a larger data set:

* `diamonds_premium_and_cheap` has `r nrow(diamonds_premium_and_cheap)` rows, while
* `diamonds_premium_or_cheap` has `r nrow(diamonds_premium_or_cheap)` rows.

```{r str comparison, tidy = FALSE}
str(diamonds_premium_and_cheap)
str(diamonds_premium_or_cheap)
```

There is, in principle, no limitation to the combination of these so-called Boolean (or logical) operators 

* and: `&` 
* or: `|` 
* equal to: `==` 
* not equal to: `!=` 
* greater than (or equal): `>` (`>=`)
* less than (or equal): `<` (`<=`)

You probably get the idea...


## Aggregating Data

Suppose we wanted to calculate the average price of the diamonds for each level of ```cut```, i.e. the average price for all diamonds of "Ideal" quality, for all diamonds of "Premium" quality and so on, this could be done using `aggregate()` like this:

```{r aggregate price cut, tidy = FALSE}
ave_price_cut <- aggregate(diamonds$price, by = list(diamonds$cut), 
                           FUN = mean)
ave_price_cut
```

Note that `by = ...` needs a list of grouping variables, even if there is only one entry. Unless a named list is specified (see 'Value' in `?aggregate`), the original column names are not carried over to the newly created table of averaged values. Instead, these get the generic names ```Group.1``` and ```x```.

```Group.1``` already indicates that we are not limited to aggregate just over one factorial variable, more are also possible. Furthermore, any  function to compute the summary statistics which can be applied to all data subsets is allowed, e.g. to compute the number of items per category we could use ```length()```:

```{r aggregate price cut + color, tidy = FALSE}
ave_n_cut_color <- aggregate(diamonds$price, 
                             by = list(diamonds$cut,
                                       diamonds$color), 
                             FUN = length)
ave_n_cut_color
```

Given that as a result of aggregating this way we loose our variable names, it makes sense to set them afterwards, so that we can easily refer to them later on.

```{r set names, tidy = FALSE}
names(ave_n_cut_color) <- c("cut", "color", "n")
str(ave_n_cut_color)
```

So, hopefully you see how useful ```aggregate()``` is for calculating summary statistics of your data.

## Sorting Data

Sorting our data according to one (or more) of the variables can also come in very handy and can be achieved using ```sort()```.

```{r sort vector, tidy = FALSE}
sort(ave_n_cut_color$n)
```

Sorting an entire data frame is a little less straightforward and can be done using ```order()```.

* for sorting according to one variable

```{r sort df, tidy = FALSE}
ave_n_cut_color <- ave_n_cut_color[order(ave_n_cut_color$cut), ]
ave_n_cut_color
```

* for sorting according to two variables

```{r sort df2, tidy = FALSE}
ave_n_cut_color <- ave_n_cut_color[order(ave_n_cut_color$cut,
                                         ave_n_cut_color$n), ]
ave_n_cut_color
```


## Merging Data

Often enough we end up with multiple data sets on our hard drive that contain useful data for the same analysis. In this case we might want to amalgamate our data sets so that we have all the data in one set.   
R provides a function called ```merge()``` that does just that:

```{r merge 2 dfs, tidy = FALSE}
ave_n_cut_color_price <- merge(ave_n_cut_color, ave_price_cut, 
                               by.x = "cut", by.y = "Group.1")
ave_n_cut_color_price
```

As the variable names of our two data sets differ, we need to specifically provide the names for each by which the merging should be done (```by.x``` and ```by.y```). The default of ```merge()``` tries to find variable names which are identical.

Note, in order to merge more than two data frames at a time, we need to call a powerful higher-order function called ```Reduce()```. This is one mighty function for doing all sorts of things iteratively.

```{r merge multiple dfs, tidy = FALSE}
names(ave_price_cut) <- c("cut", "price")

set.seed(12)

df3 <- data.frame(cut = ave_price_cut$cut,
                  var1 = rnorm(nrow(ave_price_cut), 10, 2),
                  var2 = rnorm(nrow(ave_price_cut), 100, 20))

ave_n_cut_color_price <- Reduce(function(...) merge(..., all=T), 
                                list(ave_n_cut_color, 
                                     ave_price_cut,
                                     df3))
ave_n_cut_color_price
```

Obviously, setting proper names would be the next step now...

Okay, so now we have a few tools at hand to manipulate our data in a way that we should be able to produce some meaningful graphs which tell the story that we want to be heard, or better, seen...

So, let's start plotting stuff!